---
title: "stackex"
output: html_document
date: "2024-03-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Loading the data:
```{r, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyverse)
library(effects)
library(coefplot)
library(multcomp)
library(DHARMa)
library(lme4)

size <- read_csv() #Please insert size2.csv here
size
```


Removing some columns:
```{r, include=FALSE}
size1 = subset(size,select = -c(Patch_no, 
                                Area_m2, 
                                Total_hab_type_in_map_ha, 
                                Total_hab_type_in_metric_ha, 
                                `Match_to_pre/post`, 
                                `Distance to nearest recorded broad hab type (post-only)`,
                                Notes))
size1
str(size1)
```

Looking at values in habitat_type:
```{r, include=FALSE}
unique(size1[c("Patch_hab_type")])
#There are no duplicate columns which is good
unique(size1[c("Pre_post")])
#There are NAs here so I need to remove them
```


Removing NAs in Pre_postdev:
```{r, include=FALSE}
missing <- is.na(size1$Pre_post)
sum(missing)
#There is one NA, so let me remove this
size2 <- subset(size1, subset = !missing)
dim(size1)
dim(size2)
#Successfully removed
```


# Q1: Are post-development patches smaller than pre-development patches?
Hypothesis: post-development patches are smaller than pre-development patches.  

I want to investigate whether there is a significant difference between the size of ‘pre-development’ patches and ‘post-development’ patches. Within this test, I need to account for ‘Patch habitat type’, because the average size of certain habitats are different (eg. I may find a 1 hectare grassland but not 1 hectare of pond), and certain habitats may be more common in the post-development scenario. My data are also not randomly sampled as I randomly sampled developments but then systematically sampled habitat patches within those developments. Thus, I need to do either:  
* A mixed effects model with a random effect/ grouping factor of development ID.  
* A generalised linear mixed effect model with a random effect/ grouping factor of development ID.  

Removing outliers and transforming data
```{r, include=FALSE}
#Removing the outliers
size_quartiles <- quantile(size2$Area_ha, probs=c(.25, .75), na.rm = TRUE)
size_quartiles

size_IQR <- IQR(size2$Area_ha, na.rm = TRUE)

size_Lower <- size_quartiles[1] - 1.5*size_IQR
size_Upper <- size_quartiles[2] + 1.5*size_IQR

size3 <- subset(size2, size2$Area_ha > size_Lower & size2$Area_ha < size_Upper)

```

Sorting stuff out on both the data with outliers (size2) and the data without outliers (size3):
```{r, include=FALSE}
library(lme4)

#First of all I need to change numerical to categorical value for project_ID as this is my categorical random factor. 
size3$project_ID <- as.factor(size3$project_ID)
str(size3)
#This shows me that project_ID is now a factor! Yay!
size2$project_ID <- as.factor(size2$project_ID) #Doing same thing for size2 (with outliers)
str(size2)

#I'll check the levels of this factor...
levels(size3$project_ID)
#No duplicates, yay
levels(size2$project_ID) #Doing same thing for size2 (with outliers)

#Also doing the same for habitat type:
size3$Patch_hab_type <- as.factor(size3$Patch_hab_type)
str(size3)
levels(size3$Patch_hab_type)
#Yay! no duplicates!
size2$Patch_hab_type <- as.factor(size2$Patch_hab_type) #Doing same thing for size2 (with outliers)
str(size2)
levels(size2$Patch_hab_type)

#And broad habitat type:
size3$Broad_hab <- as.factor(size3$Broad_hab)
str(size3)
levels(size3$Broad_hab)

size2$Broad_hab <- as.factor(size2$Broad_hab) #Doing same thing for size2 (with outliers)
str(size2)
levels(size2$Broad_hab)
```

Changing the packages so they've installed correctly would involve:
```{r, eval=FALSE}
o <- options(repos = "https://cran.r-project.org/")
install.packages("Matrix")
install.packages("lme4")
options(o)
```


#Plotting my data
```{r}
size3$Pre_post <- factor(size3$Pre_post, levels = c("pre", "post")) #Reorder the levels
size2$Pre_post <- factor(size2$Pre_post, levels = c("pre", "post")) #Same for size2

#Plotting size3
ggplot(size3, aes(x = Pre_post, y = Area_ha)) +
  geom_boxplot() +
  geom_point(position = "jitter", size = 2, alpha = 0.5) +
  labs(x = "Pre_post", y = "Area_ha", color = "project_ID") +
  theme_minimal()

#Plotting size2
ggplot(size2, aes(x = Pre_post, y = Area_ha)) +
  geom_boxplot() +
  geom_point(position = "jitter", size = 2, alpha = 0.5) +
  labs(x = "Pre_post", y = "Area_ha", color = "project_ID") +
  theme_minimal()
```

Trying to graph it to see the spread of other bits:
```{r}
library(viridis)
 
#By broad habitat type
ggplot(size3, aes(x = Pre_post, y = Area_ha)) +
  geom_boxplot() +
  geom_point(aes(color=Broad_hab), position = "jitter", size = 2, alpha = 0.5) +
  labs(x = "Pre_post", y = "Area_ha", color = "Broad habitat type") +
  theme_minimal()
  

#By development ID
ggplot(size3, aes(x = Pre_post, y = Area_ha)) +
  geom_boxplot() +
  geom_point(aes(color=project_ID), position = "jitter", size = 2, alpha = 0.5) +
  labs(x = "Pre_post", y = "Area_ha", color = "Development ID") +
  theme_minimal()

#Checking histogram
ggplot(size3, aes(x = Area_ha)) + geom_histogram() 
#right skewed
```



### Fitting a GLMM
My data is not normally distributed, so instead of transforming it I could alternatively use a GLMM which allows me to accommodate non-normally distributed data.

I can’t use a Mann-Whitney because my data aren’t independently sampled; I independently sampled the developments, but then within that, systematically sampled each habitat patch. 
I can’t use a Wilcoxon Sign Test because my data aren’t independently sampled and it requires paired data, which I don’t have.


My data are very right skewed so I should either use an inverse Gaussian or a Gamma distribution. These are good for right skewed data that is between 0 and positive infinity. 

Creating a GLMM with gamma distribution (without outliers):

First, trying with only project as a random effect:
```{r}
#checking how many projects i have to check the random effect:
unique_count <- length(unique(size3$project_ID))
unique_count
#There are 25 projects

#Including only project as a random effect:
glmm_gamma <- glmer(Area_ha ~ Pre_post + (1 | project_ID), 
                    family = Gamma(link = "log"), 
                    data = size3)
summary(glmm_gamma)


#This gave me the error "Warning: Model failed to converge with max|grad| = 0.0167468 (tol = 0.002, component 1)" so I'm going to try and rescale and centre my stuff:

mu <- mean(size3$Area_ha, na.rm = TRUE)
sigma <- sd(size3$Area_ha, na.rm = TRUE)
size3$Area_ha_scaled <- (size3$Area_ha - mu) / sigma
#But no I can't do that because then it's negative!

#Trying to check singularity
tt <- getME(glmm_gamma,"theta")
ll <- getME(glmm_gamma,"lower")
min(tt[ll==0])
#0.8909557, so no worry of singularity

#Trying with more iterations
glmm_gamma <- glmer(Area_ha ~ Pre_post + (1 | project_ID), 
                    family = Gamma(link = "log"), 
                    data = size3,
                    control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1000000)))
#Still didn't work

#Seeing if project_ID has an impact:
glm_no_random <- glm(Area_ha ~ Pre_post, 
                        family = Gamma(link = "log"), 
                        data = size3)
summary(glm_no_random)
#But I really can't do this because i need to account for the grouping factor

```


